---
title: Ecological interventions to prevent and manage pathogen spillover - Supplemental
  Information
author: Sokolow, SH, N Nova, K Pepin, AJ Peel, K Manlove, PC Cross, DJ Becker, RK
  Plowright, JRC Pulliam, H McCallum, and GA De Leo
date: "March 25, 2019"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    highlight: tango
    latex_engine: xelatex
  word_document: default
mainfont: Arial
geometry: margin=.75in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install these packages from CRAN if you do no have them
library(pomp)
library(readr)
library(tidyverse)
library(gridExtra)
# this package is not on CRAN; if you don't have it, you can comment out lines 13 and 35 in plotFunctions.R
library(dviz.supp) # required to call theme_dviz_hgrid()
# load external functions and parameter values
source('helperFunctions.R')
source('plotFunctions.R')
pars <- read_csv('parameters.csv')
# Format Table S1
tab1 <- pars %>% select(Parameter, Description, Name, Units, Example1, Example2) %>% mutate(Example1 = as.character(ifelse(Example1 < 0.01,signif(Example1,2),round(Example1,2))), Example2 = as.character(ifelse(Example2 < 0.01,signif(Example2,2),round(Example2,2)))) %>% rename(`Example 1` = Example1, `Example 2` = Example2)
# Format Table S2
tab2 <- pars %>% filter(!is.na(Intervention)) %>% select(Intervention, Type, Description)
```

## Modelling methods

We defined a simple 2-host system, i.e., donors and recipients, with three compartmental classes, Susceptible-Infected-Recovered (SIR), for each host species, and spillover from donors to recipients. Model simulations are conducted using an Euler-multinomial approximation to the two-host ordinary differential equation model [1]. Our model builds off frameworks previously proposed, e.g. [2, 3], but our focus is more on the practical implications of potential interventions. Using a tractable framework, we focus on the comparative outcomes of simulated management options applied to either donor or recipient populations, and highlight potential non-linearities in spillover risk that result.

We use the model to simulate disease dynamics for two sets of fixed parameter values (Table 1). We examine how each particular ecological intervention applied to a single parameter (process) affects disease outcomes in recipient populations in terms of:  1) total number of cases in the recipient population, and 2) the total number of spillover events in a defined timeframe. 

Interventions were implemented as a fixed proportional reduction in a parameter value, except donor culling and treatment are proportional increases in mortality rate ($\delta_d$) and recovery rate ($\gamma_d$ or $\gamma_r$), respectively. We assumed that each specified intervention only affected a single parameter, except vaccination of recipient, which affects both spillover transmission (donor to recipient) and transmission among recipient hosts.

```{r initialize, echo=FALSE}
# Define control parameters
SEED <- 20190326
REPS <- 100
TIMESTEP <- 7
YEARS <- 5
INITTYPE <- 'EE'
SIMID <- 8

# Define model parameters
ex1 <- pars$Example1
names(ex1) <- pars$Name
ex2 <- pars$Example2
names(ex2) <- pars$Name

propChange <- c(10^(-5:-2),0.5,seq(0.1,1,0.1))
interventions <- tab2$Type

# Initialize
inits1 <- initSim(ex1,INITTYPE)
inits2 <- initSim(ex2,INITTYPE)

fn <- paste0('simDat_',INITTYPE,'.Rdata')
fn2 <- paste0('simDatIntervention_',INITTYPE,'.Rdata')
```

Each simulation was run for `r YEARS` years using weekly time steps with initial population sizes of `r as.character(ex1[['pop0_d']])` hosts in each of the donor and recipient populations. We initialized all simulations at the endemic equilibrium (with values rounded) for the donor and at the disease-free equilibrium for the recipeint. We used parameter values representing two different example spillover systems, which differed in their assumed contact rates and durations of infection. We ran `r REPS` replicate simulations per ecological intervention condition. We presented the average outcomes (total cases in the recipient population and total number of spillover events) in Figures S3-S6. 

Model specification and assumptions are described below. Note that the modeling framework is intentionally simplistic because our goal is to visualize potential non-linearities in effects of different ecological interventions. This type of framework could be adapted to address ecological complexities of particular systems (e.g., environmental transmission, spatial structure, etc.). Currently, the framework is intended to generate hypotheses for further examination.

### _Assumptions_

- Only spillover from donor to recipient, no spillback from recipient to donor
- Homogenous mixing in each population and between them
- Direct contact transmission only
- Lifelong immunity
- No disease-induced mortality
- All newborns are susceptible
- No spatial structure
- No density-dependent processes

### _Model structure_

We first specify the ordinary differential equation (ODE) model, based on the classic SIR compartmental framework (e.g., Keeling and Rohani 2007). We then implement a stochastic, discrete time approximation to the ODE system using an Euler-multinomial approach. Disease dynamics in the donor host ($d$) are described by the following equations:

$$
\frac{dS_d}{dt} = \mu_d N_d - \delta_d S_d - \alpha_{d-d} \rho_{d-d} I_d S_d - \nu_d S_d\\
\frac{dI_d}{dt} = \alpha_{d-d} \rho_{d-d} I_d S_d - (\delta_d + \gamma_d) I_d\\
\frac{dR_d}{dt} =  \gamma_d I_d + \nu_d S_d - \delta_d R_d
$$
and in the recipient host ($r$):

$$
\frac{dS_r}{dt} = \mu_r N_r - \delta_r S_r - (\alpha_{r-r} \rho_{r-r} I_r + \alpha_{d-r} \rho_{d-r} I_d) S_r - \nu_r S_r\\
\frac{dI_r}{dt} = (\alpha_{r-r} \rho_{r-r} I_r + \alpha_{d-r} \rho_{d-r} I_d) S_r - (\delta_r + \gamma_r) I_r\\
\frac{dR_r}{dt} =  \gamma_r I_r + \nu_r S_r - \delta_r R_r
$$

where all parameters are defined as in Table S1. For simulations wtih vaccination ($\nu_d > 0$ or $\nu_r > 0$), the vaccination hazards (daily rates) are calculated from weekly vaccination probabilities $0 < p_i < 1$ as $\nu_i = -ln(1-p_i)/{\Delta t}$, where $\Delta t = 7$.

For this system, we can derive reproduction ratios $R_0$ for sub-component models (assuming no vaccination, i.e., $\nu_d = \nu_r = 0$), namely:

$$
R_{0_{d-d}} = \frac{\alpha_{d-d} \rho_{d-d} N_{0_d}}{\delta_d + \gamma_d}\\
R_{0_{d-r}} = \frac{\alpha_{d-r} \rho_{d-r} N_{0_r}}{\delta_d + \gamma_d}\\
R_{0_{r-r}} = \frac{\alpha_{r-r} \rho_{r-r} N_{0_r}}{\delta_r + \gamma_r}
$$

The disease free equilibrium for the donor in the determinisitc model, which is used to set initial conditions for this population, is:

$$
S^* = (\delta_d + \gamma_d) / (\alpha_{d-d} \rho_{d-d})\\
I^* = (\mu_d * N_{0_d} / S^* - \delta_d) / (\alpha_{d-d} \rho_{d-d})\\
R^* = \gamma_d I^* / \delta_d
$$

#### _Parameter definitions_

Table S1 shows all parameter definitions, including both model notation and variable names used in the code for model implementation. Baseline parameter values are given for two example pathogens. Example 1 represents a pathogen that has supercritical transmission in the recipient host ($R_{0_{r-r}} > 1$), like Ebola. Example 2 represents a pathogen that has subcritical transmission in the recipient host ($R_{0_{r-r}} < 1$), like Nipah virus. In both examples, the donor and recipient life expectancies are set at 15 and 60 years, respectively, and birth rates are set to balance mortality rates ($\nu_i = \delta_i$).

`r knitr::kable(tab1)`

#### _Interventions_

Table S2 shows how interventions relate to the model parameters. The `Type` column indicates how the intervention is referred to in the code.

`r knitr::kable(tab2)`

### Model implementation

All code, simulation output, and other materials necessary to reproduce this file and results presented in the main text are provided at <https://github.com/jrcpulliam/spilloverIntervention>. The Euler-multinomial approximation was implemented in R version 3.1.5 and uses the `pomp` package (version 1.19). A single update is accomplished using the following function:

```{r simEulerstep, echo=TRUE}
# Single step for Euler-multinomial implementation of model
simEulerstep <- function (x, params, dt){
  with(c(as.list(x),params),{
    N_d <- S_d + I_d + R_d
    N_r <- S_r + I_r + R_r
    beta_dd <- cont_dd * trans_dd 
    beta_dr <- cont_dr * trans_dr 
    beta_rr <- cont_rr * trans_rr 
    dFOI <- beta_dd * I_d # force of infection experienced by donor
    sFOI <- beta_dr * I_d # force of infection experienced by recipient from donor
    rFOI <- beta_rr * I_r # force of infection experienced by recipient from recipient 
    
    births_d <- rpois(n=1,lambda=birth_d*N_d*dt)
    births_r <- rpois(n=1,lambda=birth_r*N_r*dt)
    S_d.removal <- reulermultinom(n=1,size=S_d,rate=c(mort_d,dFOI,vax_d),dt=dt)
    I_d.removal <- reulermultinom(n=1,size=I_d,rate=c(mort_d,1/dur_d),dt=dt)
    R_d.removal <- reulermultinom(n=1,size=R_d,rate=c(mort_d),dt=dt)
    S_r.removal <- reulermultinom(n=1,size=S_r,rate=c(mort_r,sFOI,rFOI,vax_r),dt=dt)
    I_r.removal <- reulermultinom(n=1,size=I_r,rate=c(mort_r,1/dur_r),dt=dt)
    R_r.removal <- reulermultinom(n=1,size=R_r,rate=c(mort_r),dt=dt)
    
    # vector of changes
    c(
      dt, # change in time
      births_d-sum(S_d.removal), # change in S_d
      S_d.removal[2]-sum(I_d.removal), # change in I_d
      I_d.removal[2]-R_d.removal, # change in R_d
      births_r-sum(S_r.removal), # change in S_r
      S_r.removal[2]+S_r.removal[3]-sum(I_r.removal), # change in I_r
      I_r.removal[2]-R_r.removal, # change in R_r
      S_r.removal[2]+S_r.removal[3], # change in cum_I_r (total infections in recipient)
      S_r.removal[2] # change in cum_I_sp (total spillover infections)
    )
  })
}
```

A full realization is accomplished using the following function:

```{r runSim, echo=TRUE}
# Run a full realization for specified parameters and initial conditions
runSim <- function(init, pars, maxtime = round(YEARS*365.25), dt = TIMESTEP, browse = F){
  ts <- NULL
  pop <- init
  if(browse) browser()
  for(tt in seq(0,maxtime,dt)){
    ts <- rbind(ts,pop)
    pop <- pop + simEulerstep(pop,pars,dt)
  }
  return(data.frame(ts))
}
```

```{r baseline, echo=FALSE}
if(file.exists(fn)){
  load(fn)
}else{
  set.seed(SEED)
  sim1 <- replicate(REPS,runSim(inits1,ex1))
  sim2 <- replicate(REPS,runSim(inits2,ex2))
  save(sim1,sim2,file = fn)
}
```

### Model output

#### _Example trajectories_

_Example 1_:

```{r simplot1, echo=FALSE}
grid.arrange(simPlot(sim1[,SIMID]),foiPlot(sim1[,SIMID],ex1), nrow = 2)
```

_Example 2_:

```{r simplot2, echo=FALSE}
grid.arrange(simPlot(sim2[,SIMID]),foiPlot(sim2[,SIMID],ex2), nrow = 2)

# sapply(1:REPS,function(ii) sim2[,ii]$I_d[261])
```

#### _Impact of interventions_

```{r interventions, echo=FALSE}
if(file.exists(fn2)){
  load(fn2)
}else{
  set.seed(SEED)
  tmp <- expand.grid(prop = propChange, intv = interventions, baseline = c('ex1','ex2'))
  tmp$intv <- as.character(tmp$intv)
  tmp$baseline <- as.character(tmp$baseline)
  res <- NULL
  for(ii in 1:nrow(tmp)){
    res <- rbind(res,simSummary(tmp$prop[ii],tmp$intv[ii],tmp$baseline[ii],nn=REPS))
  }
  save(res,file = fn2)
}

```

_Example 1_:

```{r ex1, echo=FALSE}
resPlot_all('ex1')
resPlot_sp('ex1')
```

_Example 2_:

```{r ex2, echo=FALSE}
resPlot_all('ex2')
resPlot_sp('ex2')
```